Movement in Simulations over Time

### 2025-03-20 sim_movement_over_time.qmd
more investiagtions to network structure

```{python}
import h5py
import polars as pl
import numpy as np
from matplotlib import pyplot as plt

@@

```{python}
import igraph as ig

@@

static_df = pl.from_dict(
    {
        'indirect': S.es['indirect_weight'],
        # 'link_time': S.es['link_time'],
        'source': S.es['source_loc'],
        'target': S.es['dest_loc'],
    }
).cast(pl.Int64)

@@

dep_rates = pl.concat([
    snapshot_df
        .with_columns(model=pl.lit('snapshot'))
        .select('dep_rate', 'source', 'target', 'model'), 
    static_df
        .with_columns(model=pl.lit('static'))
        .select('dep_rate', 'source', 'target', 'model')
])

```
```{python}
hist_opts = {
    'histtype': 'step',
    'bins': 31,
    'density': True
}
for name, arr in zip(['snapshot', 'static'], [snapshot_moves, static_moves]):
    plt.hist(arr, **hist_opts, label=name)

plt.yscale('log')
plt.legend()

>>>>

```{python}
import h5py
import polars as pl
import numpy as np
from matplotlib import pyplot as plt
import seaborn as sns

@@

```{python}
import igraph as ig
from collections import Counter

@@

static_df = pl.from_dict(
    {
        'indirect': S.es['indirect_weight'],
        'direct': S.es['direct_weight'],
        'source': S.es['source_loc'],
        'target': S.es['dest_loc'],
    }
).cast(pl.Int64)

@@

dep_rates = pl.concat([
    snapshot_df
        .with_columns(model=pl.lit('snapshot'))
        .select('dep_rate', 'source', 'target', 'model'), 
    static_df
        .with_columns(model=pl.lit('static'))
        .select('dep_rate', 'source', 'target', 'model')
])
```

```{python}
sns.histplot(
    dep_rates.group_by('source', 'model').agg(pl.col('dep_rate').sum()),
    x='dep_rate',
    hue='model',
)
```

I have a hunch about this. Let's construct the "combined" graph of the snapshots over time, and see when it becomes conencted.
This will be the minimum time before we can hit equilibirum.
```{python}
NO_CACHE = False
connectedness_file = "conc_tempo_14_detailed/analysis/overlay_connectedness.parquet"

if NO_CACHE:
    Gs = [
        ig.Graph.Read(f"conc_tempo_14_detailed/{i:04d}.graphml")
        for i in range(0, 3626+14, 14)
    ]

    # remove problem vertex (isolated hospital)
    problem_vertex = 4454
    for graph in Gs:
        graph.delete_vertices(graph.vs.select(name_eq=problem_vertex).indices)

    connectedness = []
    cluster_sizes = []
    for i in range(len(Gs)):
        U = ig.union(Gs[:i+1])
        connectedness.append(U.is_connected())
        cluster_sizes.append(Counter(map(len, U.connected_components())))

    df = pl.from_dicts([
        {'snapshot': i, 'fully_connected': full_conn, 'cluster_size': k, 'count': v} 
        for i, (full_conn, clus_counter) in enumerate(zip(connectedness, cluster_sizes)) 
        for k,v in clus_counter.items()
    ])

    df.write_parquet(connectedness_file)
else:
    df = pl.read_parquet(connectedness_file)
```

```{python}
fig, axs = plt.subplot_mosaic([['c'], ['c'], ['b']], sharex=True, layout='constrained')

cluster_stats = df.group_by('snapshot').agg(
    pl.col('cluster_size').max().alias('dominant_cluster_size'),
    pl.col('count').sum().alias('number_of_clusters')
).with_columns(time = pl.col('snapshot') * 14).sort('snapshot')

sns.lineplot(cluster_stats, x='time', y='dominant_cluster_size', ax=axs['c'])
sns.lineplot(cluster_stats, x='time', y='number_of_clusters', ax=axs['b'])
```

## Rare path exploration

We want to see if we can come up with a way of detecting and analysing "rare" paths.
Specifically rare 1-paths (potentially of the indirect flavour).

To do this, we will want to tabulate edges of the snapshot against the static edge weights

```{python}
# static ref table already extracted in earlier cell

NO_CACHE = False

snapshot_table_file = "conc_tempo_14_detailed/analysis/combined_outwards.parquet"

if NO_CACHE:
    snapshot_graph_tables = []
    for i in range(0, 3626+14, 14):
        graph = ig.Graph.Read(f"conc_tempo_14_detailed/{i:04d}.graphml")
        problem_vertex = 4454
        graph.delete_vertices(graph.vs.select(name_eq=problem_vertex).indices)
        snapshot_graph_tables.append(
            (pl.from_dict({
                'source': graph.es['source'],
                'target': graph.es['target'],
                'snapshot_direct': graph.es['weight'],
                'snapshot_indirect': graph.es['departures'],
            })
            .cast({
                'source': pl.Int64,
                'target': pl.Int64,
                'snapshot_direct': pl.Int64,
                'snapshot_indirect': pl.Int64,
            })
            .with_columns(time=pl.lit(i))
            )
        )
    snapshot_table = pl.concat(snapshot_graph_tables)
    snapshot_table.write_parquet(snapshot_table_file)
else:
    snapshot_table = pl.read_parquet(snapshot_table_file)

joined_table = snapshot_table.join(static_df, on=['source', 'target'])

SS_WIDTH = 14

joined_table = (
    joined_table
    .with_columns(
        ss_direct_rate = pl.col('snapshot_direct').cast(pl.Float64) / pl.col('estimated_size') / SS_WIDTH,
        ss_indirect_rate = pl.col('snapshot_indirect').cast(pl.Float64) / pl.col('estimated_size') / SS_WIDTH,
        st_direct_rate = pl.col('direct').cast(pl.Float64) / pl.col('estimated_size') / S['time_span'],
    )
)
```

```{python}
devs = joined_table.filter(pl.col('dep_rate') > 0.0).select(pl.col('ss_indirect_rate') - pl.col('dep_rate')).to_series().to_numpy()
plt.hist(devs, bins=301)
plt.yscale('log')
plt.axvline(3 * np.sqrt(np.mean(devs**2)), color='red')
plt.axvline(-3*np.sqrt(np.mean(devs**2)), color='red')
```
